---
title: "Practical Machine Learning Course Project"
author: "Soni Baidya"
date: "July 6, 2016"
output: html_document
---

# Summary

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, our goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.


```{r echo =FALSE, message=FALSE, warning=FALSE}
library(data.table)
library(caret)
library(randomForest)
library(foreach)
library(rpart)
library(rpart.plot)
library(corrplot)
```

# Read the data

Read both training and testing data which are in the current folder.

```{r}
training_data <- read.csv("pml-training.csv", na.strings=c("#DIV/0!"," ", "", "NA", "NAs", "NULL"))
testing_data <- read.csv("pml-testing.csv", na.strings=c("#DIV/0!"," ", "", "NA", "NAs", "NULL"))
```

# Clean the data
We need to drop columns with NAs, drop highly correlated variables and drop variables with 0 (or approx to 0) variance. The results are hidden as they take a very long space.

```{r}
str(training_data)
cleantraining <- training_data[, -which(names(training_data) %in% c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "new_window", "num_window"))]
# This drops columns with NAs
cleantraining = cleantraining[, colSums(is.na(cleantraining)) == 0] 
zerovariance =nearZeroVar(cleantraining[sapply(cleantraining, is.numeric)], saveMetrics=TRUE)
# To remove 0 or near to 0 variance variables
cleantraining = cleantraining[, zerovariance[, 'nzv'] == 0]
correlationmatrix <- cor(na.omit(cleantraining[sapply(cleantraining, is.numeric)]))
dim(correlationmatrix)
correlationmatrixdegreesoffreedom <- expand.grid(row = 1:52, col = 1:52)
# This returns the correlation matrix in matrix format
correlationmatrixdegreesoffreedom$correlation <- as.vector(correlationmatrix) 
removehighcorrelation <- findCorrelation(correlationmatrix, cutoff = .7, verbose = TRUE)
# This removes highly correlated variables (in psychometric theory .7+ correlation is a high correlation)
cleantraining <- cleantraining[, -removehighcorrelation] 

for(i in c(8:ncol(cleantraining)-1)) {cleantraining[,i] = as.numeric(as.character(cleantraining[,i]))}
# Some columns were blank, hence are dropped. I will use a set that only includes complete columns. I also remove user name, timestamps and windows to have a light data set.
for(i in c(8:ncol(testing_data)-1)) {testing_data[,i] = as.numeric(as.character(testing_data[,i]))} 

featureset <- colnames(cleantraining[colSums(is.na(cleantraining)) == 0])[-(1:7)]
modeldata <- cleantraining[featureset]
#now we have the model data built from our feature set.
featureset 
```

# Model
We need to split the sample in two samples. 60% for training and 40% for testing is the usual.

```{r}
idx <- createDataPartition(modeldata$classe, p=0.6, list=FALSE )
training <- modeldata[idx,]
testing <- modeldata[-idx,]
```

A predictive model is fitted using Random Forest algorithm. Highly correlated variables were already removed but still this algorithm is robust to correlated covariates and outliers. A 5 fold cross validation is used.

```{r}
control <- trainControl(method="cv", 5)
model <- train(classe ~ ., data=training, method="rf", trControl=control, ntree=250)
model
```

The performance of the model is estimated on the validation data set.  

```{r}
predict <- predict(model, testing)
confusionMatrix(testing$classe, predict)

accuracy <- postResample(predict, testing$classe)
accuracy
```

The estimated accuracy of the model is 97.6% and the estimated out of sample error is 2.4%.

# Predictions
The model is aplied to the original testing data.

```{r}
result <- predict(model, training[, -length(names(training))])
```  

# Tree

```{r}
treeModel <- rpart(classe ~ ., data=cleantraining, method="class")
prp(treeModel) 
```

# The Question's answer

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("Prediction_file_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

testing_data <- testing_data[featureset[featureset!='classe']]
answers <- predict(model, newdata=testing_data)
answers

pml_write_files(answers)
```